\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-11 paper/examples.tex
paper: rewrote example section to reference cdc data

\section{Examples}

Consider the case of an analyst looking to explore the sources and causes of
polling error in a presidential election. One hypothesis \cite{zukin2015s}
is that the rapid switch from landlines to cellphones increases the
difficulty of contacting potential poll respondents. The analyst may
begin her analysis by comparing the polling error in a state with the
distribution of households with landlines, for the set of states for which
polling error was most pronounced. In order to probe deeply, the analyst
accesses a household-level survey that includes data on
whether the household has a landline. One such survey --- which will be
discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
conducted by the US Census Bureau.

Surveys are rife with missing values due to non-response and other issues, and
the ACS is no exception. The Census Bureau and other similar organization attempt to present
data in as unadulterated a fashion as possible. The ACS dataset is not amenable to one-time
imputation by either the Census Bureau, for this reason, or the analyst's organization,
which likely uses the same database for a variety of unrelated uses (asking for different
imputation strategies). Thus, the analyst must deal with missing values in one way or
another.

In this case, a non-negligible fraction of respondents omitted information on whether they
have a landline. The analyst hypothesizes that older houses are more likely to have
landlines and that households that are relatively new are less likely to have landlines. 
The analyst could take advantage of these presumed correlations to impute the missing values
on the base table in one expensive operation. On the other hand, the query could be run
directly and rows with missing values may be dropped entirely. With ImputeDB, the analyst
could impute the relevant subset of data on-the-fly, achieving better performance without
losing the information in the dropped examples.

The analyst submits a query (Figure~\ref{fig:example-query}) to ImputeDB which finds an optimized query
plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
emphasis parameter, in order to receive a plan that does an expensive imputation step. See
Section~\ref{sec:cost-model} for more details.}. The resulting query plan
(Figure~\ref{fig:query-plan})
imputes missing values for \verb!acs.TEL! after states with high polling error have been
selected. This decision reduces the set of tuples that must be input into the imputation
algorithm to just those that are relevant to the query. Later, we show the algorithms that
the ImputeDB query optimizer uses in order to detect these opportunities and control the
tradeoffs between efficiency and imputation quality. Finally, note that the optimizer doesn't
place any imputation operator for the \verb|acs.ST| column, as it uses a histogram to
detect that there are no missing values there (as might be expected for a survey question that isn't
impacted by non-responses).

\begin{figure}
\begin{lstlisting}[language=SQL]
SELECT polling.ST, AVG(acs.TEL)
FROM polling, acs
WHERE polling.ST = acs.ST
  AND polling.ERROR > 50         -- 5 percentage points
GROUP BY polling.ST;
\end{lstlisting}
\caption{A typical analyst query on ACS data}
\label{fig:example-query}
\end{figure}

\begin{figure}[!ht]
    \centering
    \includegraphics[width=0.6\textwidth]{figures/example.png}
    \caption{Query plan generated by ImputeDB prioritizing imputation quality}
    \label{fig:query-plan}
\end{figure}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

\section{Motivating Example}
Consider the case of an epidemiologist tasked with an initial
exploratory analysis of data collected for individuals across
a battery of exams. In particular, she is interested in exploring
the relationship between income and the immune system,
controlling for factors such as weight and gender.

While careful modeling techniques will be necessary before putting
results into the final research paper, the epidemiologist is anxious
to get a quick and relatively accurate view into the data. However,
the data has been collected through surveys, namely CDC surveys (see Section~\ref{subsec:datasets} for details on such data),
and there is a significant amount of missing data across all
relevant attributes. The researcher must know develop a strategy
for filling in such missing values. She could drop any records with
missing values and continue her analysis with the remaining data,
or she could take advantage of existing correlations between attributes
to fit a more sophisticated model on the base tables, which may take
a long time.

To complicate matters further, the epidemiologist realizes that
various of the steps she is interested in, including filtering and grouping,
are impacted differently by missing data, and that these step may change
repeatedly, as she wants to run various queries during the exploration phase.

Rather than work through
the implications of each case, she submits a query, as shown in Figure~\ref{fig:example-query}, to ImputeDB. The system, parameterized by a value indicating
a tradeoff between information loss and runtime cost, then finds
the optimal query plan, as shown in Figure~\ref{xxx}.

Tens of queries later, the epidemiologist has the holistic view of the
data that they required before carefully crafting their own tailored
imputation model. Knowing that there may be a need to explore
this data set further, they can easily incorporate their imputation model
into ImputeDB for future use.

\begin{figure}
\begin{lstlisting}[language=SQL]
SELECT
income,
AVG(white_blood_cell_ct)
FROM demo, exams, labs
WHERE 
gender = 2 AND weight >= 120 AND
demo.id = exams.id AND exams.id = labs.id
GROUP BY demo.income
\end{lstlisting}
\caption{A typical public health query on CDC's NHANES data}
\label{fig:example-query}
\end{figure}

\todo{add a small comment about the number of plans in the plancache
to highlight size of search space}
\todo{add a figure with the resulting query plan, and what its doing}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
